Double White Dwarf Mergers on Adaptive Meshes\\ I. Methodology and Code Verification

### 2015-08-14 papers/wdmerger_I/paper.tex
more comments

We loaded a single star of mass $0.9\ \msolar$ onto the grid at the origin, 
and evolved it for 200 seconds. Our diagnostic of choice is the effective 
radius of the star, determined by the volume of the grid that has a density 
greater than $10^3\ \text{g cm}^{-3}$ (see \autoref{sec:implementation} 
for details on this measure). This choice of density is intended to 
mark a reasonable outer edge to the star that is not immediately susceptible 
to the numerical errors prevalent near the physical edge of the star. 
\autoref{fig:single_star_static_radius} shows our results at various resolutions. 
As expected, the star quickly approaches an equilibrium size that is different 
(and in this case larger) than the one-dimensional model. However, the magnitude 
of this change becomes smaller with resolution. The star is only approximately 
in equilibrium when the coarse grid of $256^3$ zones has a level of 
refinement that jumps by a factor of four. However, even then there is a slight 
uptick in the size toward the event, implying that the numerical stability 
is not guaranteed for arbitrarily long timescales. This result suggests that when 
doing the equilibrium initial models that will former the basis of later calculations,
we should carefully monitor the evolution of the stars before applying any 
artificial damping to cause the merger, to ensure that the merger is due to 
this applied force and not the intrinsic numerical instability of the stars.

@@

\begin{figure*}[ht]
  \centering
  \includegraphics[trim=0.75in 0.45in 0.85in 0.5in, clip, scale = 0.75]{{{plots/kh_t2.0_p3_low_res_collage}}}
  \caption{2D Kelvin-Helmholtz instability test at $t = 2.0$ for the initial 
    conditions given by \autoref{eq:kh_p3_rho} through \autoref{eq:kh_p3_vy}. 
    The meaning of the rows and columns is the same as in \autoref{fig:kh_p2}.
    \label{fig:kh_p3}}
\end{figure*}

@@

\citet{hopkins:2015} performed this test as part of the testing of their code GIZMO. 
They showed the late-time evolution of this system, when non-linear effects have 
taken over and significantly disrupted the initial flow. However, at low resolution the 
tested grid algorithm had failed to disrupt both for $v = 0$ and $v = 10$. We too ran 
this problem until $t = 10$, and confirm that the Kelvin-Helmholtz instability damps out 
at low resolution but goes strongly non-linear and disrupts the flow at high resolution.
However, we strongly emphasize the point that this does not objectively 
demonstrate a defiency in fixed-grid codes for this problem. We can only determine 
the validity of a method when we have a trustworthy, converged solution to compare to, and 
this is lacking for this problem at late times. As observed by \citeauthor{mcnally:2012}, 
this is because the secondary instabilities form for this problem when the whorls of 
the Kelvin-Helmholtz tendrils stretch out and create gradients that approach the 
grid resolution. This is prime breeding ground for numerical noise. But because 
the nature of this noise will depend on the resolution, it will be very different for 
simulations at different resolutions. If these instabilities are seeded because of 
this resolution-dependent noise and are not seeded instead in a controlled manner 
such that they appear at the same time and location, then we simply cannot draw 
any conclusions that bear on the question of verification from this test at late times. 
\autoref{fig:kh_p3_high_res} provides a sense of this by examining the crucial 
time at which the transition from the linear to the non-linear regime is occurring. 
At all of these very high resolutions the secondary instabilities develop, but they 
occur at different times and have different spatial scales for each resolution.

@@

To see this effect in action for a stellar simulation, we repeated the test of  
\autoref{sec:HSE} with a bulk velocity on the grid. We chose a velocity of $2.56 \times 10^{8}\ \text{cm s}^{-1}$.
For context, this is comparable to the orbital velocities of the stars in \autoref{sec:kepler}.
This test was inspired by \citet{tasker:2008}, who considered a moving galaxy cluster 
and who obtained a long timescale evolution by using periodic boundary conditions, 
so that the cluster would cross the domain multiple times throughout the evolution.
We believe that periodic boundary conditions are unrealistic for our type of simulation,
so we prefer to do one continuous simulation where the star does not cross the boundaries.
Since our normal grid was not large enough to allow the motion to continue for very long, we expanded 
the domain size by a factor of four, and then included an extra refined level around the star to 
keep the effective resolution the same. We started the star off in the lower left corner of the domain, and 
pointed its velocity towards the upper right corner. This allowed us to evolve the star for the 
same length of time as for the original test. We note that getting the gravity boundary conditions right 
required us to move the origin of the problem at the bulk velocity, so that the multipole moments 
were computed with respect to the stellar center.

@@

In \autoref{fig:single_star_compare_radius}, we take the results of 
\autoref{sec:HSE} (the ``static'' case), and plot on top of it the results of this 
new simulation (the ``motion'' case). We see immediately that this bulk velocity
causes the star to be much worse at maintaining hydrostatic equilibrium. Not only 
is the absolute size of the star significantly larger (nearly a factor of two 
at the lowest feasible resolution we consider), but also there is a clear upward 
trend in the size that has not terminated at any resolution by the end of the simulation.
This again emphasizes the results mentioned earlier, that we must be careful not 
to trust any simulation with significant mass transfer if we are not confident that the 
mass transfer is seeded in a controllable manner and free from numerical noise.

>>>>

We loaded a single star of mass $0.9\ \msolar$ onto the grid at the
origin, and evolved it for 200 seconds. Our diagnostic of choice is
the effective radius of the star, determined by the volume of the grid
that has a density greater than $10^3\ \text{g cm}^{-3}$ (see
\autoref{sec:implementation} for details on this measure). This choice
of density is intended to mark a reasonable outer edge to the star
that is not immediately susceptible to the numerical errors prevalent
near the physical edge of the star.
\autoref{fig:single_star_static_radius} shows our results at various
resolutions.  As expected, the star quickly approaches an equilibrium
size that is different (and in this case larger) than the
one-dimensional model. However, the magnitude of this change becomes
smaller with resolution. The star is only approximately in equilibrium
when the coarse grid of $256^3$ zones has a level of refinement that
jumps by a factor of four. However, even then there is a slight uptick
in the size toward the event, implying that the numerical stability is
not guaranteed for arbitrarily long timescales. This result suggests
that when doing the equilibrium initial models that will former the
basis of later calculations, we should carefully monitor the evolution
of the stars before applying any artificial damping to cause the
merger, to ensure that the merger is due to this applied force and not
the intrinsic numerical instability of the stars. \MarginPar{you never referenced figure 4}

@@

\begin{figure*}[ht]
  \centering
  \includegraphics[trim=0.75in 0.45in 0.85in 0.5in, clip, scale = 0.75]{{{plots/kh_t2.0_p3_low_res_collage}}}
  \caption{2D Kelvin-Helmholtz instability test at $t = 2.0$ for the initial 
    conditions given by \autoref{eq:kh_p3_rho} through \autoref{eq:kh_p3_vy}. 
    The meaning of the rows and columns is the same as in \autoref{fig:kh_p2}.
    \label{fig:kh_p3}  {\color{red} You should note this is $R_B$}}
\end{figure*}

@@

\citet{hopkins:2015} performed this test as part of the testing of
their code GIZMO.  They showed the late-time evolution of this system,
when non-linear effects have taken over and significantly disrupted
the initial flow. However, at low resolution the tested grid algorithm
had failed to disrupt both for $v = 0$ and $v = 10$. We too ran this
problem until $t = 10$, and confirm that the Kelvin-Helmholtz
instability damps out at low resolution but goes strongly non-linear
and disrupts the flow at high resolution.  However, we strongly
emphasize the point that this does not objectively demonstrate a
defiency in fixed-grid codes for this problem. We can only determine
the validity of a method when we have a trustworthy, converged
solution to compare to, and this is lacking for this problem at late
times. As observed by \citeauthor{mcnally:2012}, this is because the
secondary instabilities form for this problem when the whirls of the
Kelvin-Helmholtz tendrils stretch out and create gradients that
approach the grid resolution. This is prime breeding ground for
numerical noise. But because the nature of this noise will depend on
the resolution, it will be very different for simulations at different
resolutions. If these instabilities are seeded because of this
resolution-dependent noise and are not seeded instead in a controlled
manner such that they appear at the same time and location, then we
simply cannot draw any conclusions that bear on the question of
verification from this test at late times.
\autoref{fig:kh_p3_high_res} provides a sense of this by examining the
crucial time at which the transition from the linear to the non-linear
regime is occurring.  At all of these very high resolutions the
secondary instabilities develop, but they occur at different times and
have different spatial scales for each resolution.

@@

To see this effect in action for a stellar simulation, we repeated the
test of \autoref{sec:HSE} with a bulk velocity on the grid. We chose a
velocity of $2.56 \times 10^{8}\ \text{cm s}^{-1}$.  \MarginPar{Approx
  what Mach number is this?}  {For context, this is comparable to the
  orbital velocities of the stars in \autoref{sec:kepler}.  This test
  was inspired by \citet{tasker:2008}, who considered a moving galaxy
  cluster and who obtained a long timescale evolution by using
  periodic boundary conditions, so that the cluster would cross the
  domain multiple times throughout the evolution.  We believe that
  periodic boundary conditions are unrealistic for our type of
  simulation, so we prefer to do one continuous simulation where the
  star does not cross the boundaries.  Since our normal grid was not
  large enough to allow the motion to continue for very long, we
  expanded the domain size by a factor of four, and then included an
  extra refined level around the star to keep the effective resolution
  the same. We started the star off in the lower left corner of the
  domain, and pointed its velocity towards the upper right
  corner. This allowed us to evolve the star for the same length of
  time as for the original test. We note that getting the gravity
  boundary conditions right required us to move the origin of the
  problem at the bulk velocity, so that the multipole moments were
  computed with respect to the stellar center.  \MarginPar{are you
    sure that the single WD left quiet in the {\em corner} of the
    bigger domain is as stable as it was in the earlier test?}

@@

In \autoref{fig:single_star_compare_radius}, we take the results of
\autoref{sec:HSE} (the ``static'' case), and plot on top of it the
results of this new simulation (the ``motion'' case). We see
immediately that this bulk velocity causes the star to be much worse
at maintaining hydrostatic equilibrium. Not only is the absolute size
of the star significantly larger (nearly a factor of two at the lowest
feasible resolution we consider), but also there is a clear upward
trend in the size that has not terminated at any resolution by the end
of the simulation.  This again emphasizes the results mentioned
earlier, that we must be careful not to trust any simulation with
significant mass transfer if we are not confident that the mass
transfer is seeded in a controllable manner and free from numerical
noise.  \MarginPar{I really think that we should be able to do better here---this is where I'd like to see ppm\_type = 2, or better hydro-gravity coupling (eventually)}
